Non-word repetition in Yélî Dnye: A small-scale study

### 2019-10-29 manuscript.Rmd
fleshing out

- many studies on phonological development in childhood
- they show phonology continues developing as children learn more words and gain mastery of phonetic targets
- one key task in this research in NWR
- used for theory (eg what are the links between phonology, working memory, and lexicon?), and for applications (nwr as a diagnostic)
- it has been used acros a wide array of languages, particularly in Europe
- however, less is known about NWR development in non-Indo European languages
- this study: NWR in a Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory

In this study, we use nonword repetition (NWR), a technique that has been widely adopted to study the development of perception-production across multiple languages (CITE), particularly in Europe (CITE). In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning. The child should immediately try to say it back without changing anything. Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online working memory (to encode the item in the interval between hearing it and saying it back). In our implementation of NWR, we additionally made sure that some of the items contained typologically rare and/or  challenging sounds, to additionally capture children's auditory-articulatory accuracy.

@@

### Intro to the language

### Intro to the people

@@

- typical structure of NWR tasks: N of items
- complexity varies widely across tasks, sometimes within tasks on purpose
- some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
- calculation of frequency in Yêli is challenging because it is a low-resource language 
- so we only manipulated structural complexity, not probability, in the design of the stimuli
- previous work seems to avoid difficult sounds
- but we felt this was important to represent the language, so we also varied this factor

- previous work shows even 2yo can do the task
- but there are striking changes with age
- inconsistent results RE effects of SES (including parental education) & child sex
- bilingual disadvantage (could be due to lower exposure to input; learning 2 phonologies; learning 2 lexica)

- brief summary of inconsistent results SES

- idem for sex

### Research questions and predictions

- What is the accuracy (whole word, phoneme based, distance) overall?
- How does this change as a function of item complexity (nb syll, syll complexity, sound complexity)?
- How frequent are errors that result in real words? Is that a function of item complexity?
- Is individual variation explainable by child age, sex, parental education, birth order, monolingual status?

Based on previous work, we made the following predictions:

1. Children are more accurate for mono-syllables than longer items
2. Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds. Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
3. Children's accuracy increases with child age.
4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language) 
<!-- **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)** ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)-->
5. Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children

## Pre-registration and planned analyses

In view of the hypothesis driven nature of this work, we intended to boost the interpretational value of these data by announcing our analysis plans prior to conducting them. There are a few limitations in our approach:

- The sample size is too small to have enough power to detect all of the relationships of interest
- In fact, the sample size is too small to detect some of the known effects (see Table XX **add table with r/ds from previous work**)
- By virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies
- Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear **(MC: i guess this isn't true anymore given all the Anna data?)**) on the first 11 participants, and who segmented all children's production. Therefore, a key condition of blinding is not met

Nonetheless, some form of pre-registration still seemed preferrable over the complete lack of it.

The pre-registration can be found XXX

## Stimuli

Many  NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language. We kept the same variation in item length and the non-meaningfulness requirement **(MC: before i forget, note that one of the stims, 'yi', is in fact a word, and a pretty frequent one at that (meaning 'tree'); i'm not sure how that happened, sorry!)**, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language. We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.

A first list of candidate items was generated in 2018 by selecting simple consonants (LIST) and vowels (LIST) that were combined into CV syllables, further sampling the space of 1- to 4-syllable sequences. These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary. They were further presented orally to three local research assistants, who were asked to repeat them and further say whether they were real words. Any item for which two or more of the assistants reported them having a meaning or some form of association were was excluded.

A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce CV monosyllables. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out **MC: i had the impression that you relaxed this short/long constraint in the end, but maybe i dreamt that :)**. Since the phonology and phonetics of Yélî is still in the process of being described (CITE), there could have been undocumented constraints that rendered items illegal **MC: well any intervocalic d that isn't a retroflex flap would be phonotactically illegal, i think---do we have any of those? i thought we maybe did but i don't remember and i don't remember, if so, what ghaalyu pronounced in his recording**. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence). These candidates were presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded using a headset XXX and an olympus XXX from the written form presented together with the same item orally. The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.

The final list is XX.

A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets. The split had the following constraint:

- the same 3 items were selected as practice items and used in all 40 elicitation sets
- splits were done within each length group from the 2018 items (i.e., separately for 1, 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists). Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
- once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2_syllables, then 3-syllables, and finally 4-syllable items.

The 40 elicitation sets are available online from XXX.

## Procedure

@@

Indeed, we visited four different hamlets once, and attempted to test all eligible childr en present at that time, to prevent the items 'spreading' through hearsay. In the first village, we tested children in five different places, with some children being tested inside their house and others tested on the veranda of a house even if it was not their own. The complete list of places and the ways in which they met the desiderata mentioned above can be found in supplementary materials, under comments for each child (direct link xXX).

The child was donned a headset (xx for most of the children, xx for the rest). For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth. A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the 3 practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything. An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.

@@

## Coding

@@

Additionally, the experimenter listened through all recordings and segmented out every child production. For the first 11 participants, she additionally made a judgment of whether the item was correctly or incorrectly repeated, and transcribed what she heard, which provides both an accuracy measure and a qualitative description of what children were actually saying in response to each stimulus. . We call this "non-native coding," because this experimenter is not familiar with Yélî and has insufficient training on Yélî sounds, including common allophonic variants, not to mention not having some of the sounds in her native perceptual inventory.

Therefore, we sought additional information by asking a local informant to listen through each child production, paried with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated. We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets). This "native coding" provides both accuracy and qualtiative desciptions of what the child said that are likely more reliable than those of the non-native coder. **MC: if we have anna's codes for everything, do we need your coding for anything?**

## Analyses

For NWR accuracy, we considered only the child's first attempt, which was scored as correct or incorrect online, by a non-native coder (ofr the first 11 kids), and a native coder.

Some NWR studies employ phoneme accuracy (rather than whole word accuracy). We used exclusively the native coding for these, and we calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).

Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account, again using only the native coder's transcriptions.

## Participants

Total N of kids, N of families, N of villages. Some children could not be included for the following reasons: refused participation (N=2), failed to repeat items presented over headphones even after coaching (N=3), spoke too softly to allow offline coding (N=8). In addition, one teenager was tested to put children at ease; their data is not included in analyses below. The remaining XX exhibited the following demographic characteristics

INSERT TABLE WITH THE KEY VARS MENTIONED ABOVE:

- child age
- monoling
- mat ed
- birth order

@@

read.csv("NWR-demo.csv",header=T)->demo
colnames(demo)
subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc

summary(inc)

read.table("log2.txt",header=T,sep="\t")->x
x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
x[order(x$out),]->x
x$cor.online=1
for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0

@@

plot(ms$score.online~ms$score)

### Exploratory analyses: The role of phonotactic frequency

@@

@@

>>>>

Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., CITE). One common task in this line of research is nonword repetition (NWR). In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning. The child should immediately try to say it back without changing anything. Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online working memory (to encode the item in the interval between hearing it and saying it back).

NWR has been used to seek answers to theoretical questions, including what are the links between phonology, working memory, and the lexicon [CITE], as well as for applications, notably using nwr as a diagnostic for language delays and disorders [CITE]. Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe [CITE].

In this study, we use NWR to measure perception-production in children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory. In our implementation of NWR, we  made sure that some of the items contained typologically rare and/or  challenging sounds.

### Intro to the language (@mc) - please feel free to throw away anything that is not useful!

### Intro to the people (@mc) - please feel free to throw away anything that is not useful!

@@

@@

Although the procedure tends to be quite stable, there is some variation in the presentation and structure of items found in previous NWR work. For example, items are often presented orally by the experimenter [CITE], although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation [CITE]. Additionally, while some studies have used 10-15 non-words, others have employed up to 30 unique items [CITE]. Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) as well as syllable structure [open as opposed to closed syllables, CITE].

Previous work seems to avoid difficult sounds, but we felt this was important to represent the language, so we also varied this factor. We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children. This allowed us to get information about repetition accuracy of more items.

Naturally, designing the task in this way may render the study of individual variaiton within the population  more difficult because different children are exposed to different items. However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit. That said, we contribute to the literature by also reporting descriptive analyses that could potentially be integrated in meta-analytic efforts. Based on previous work, we looked at potential improvements with age [CITE], and potential negative effects of bilingual exposure [CITE]. Previous work typically finds no significant differences as a function of maternal education [see CITE but also CITE] or child gender [CITE]. Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [CITE], and therefore we report on those too.

### Research questions

We sought to address the following questions:

- What is the overall repetition accuracy (whole word, phoneme based, distance)?
- How does this change as a function of item complexity (number of syllablex, sound complexity)?
- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?

In view of the hypothesis-driven nature of this work, we consider boosting the interpretational value of these data by announcing our analysis plans prior to conducting them. However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects. For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small. **add table with r/ds from previous work**)
Moreover, by virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies. Therefore, all analyses here are descriptive and should be considered exploratory.

@ac ADD TABLE SUMMARIZING PREVIOUS WORK

- age differences
- length effects

## Methods

### Stimuli

Many  NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language. We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language. We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.

A first list of candidate items was generated in 2018 by selecting simple consonants ("p",	"t",	"d",	"k",	"m",	"n",	"w",	"y") and vowels ("i", "o","u", "a","e") that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences. These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary. MC presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words. Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.

A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce consonant-vowel monosyllables. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out. Since the phonology and phonetics of Yélî is still in the process of being described [CITE], there could have been undocumented constraints that rendered items illegal. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence). These candidates were presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded using a headset XXX and an olympus XXX from the written form presented together with the same item orally. The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words. Even so, there was one monosyllable that was often identified as a real word (intended "yî" /yXX/; identified as "yi" /yi/, *tree*). This item is removed from analyses.

@mc can you please add the IPA in // for all of the items below?
The final list is composed of three practice items (wî, poni, nopimade); 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives (
dpa
dp:a
dpâ
dpéé
dpê
dpi
dpu
kp:ââ
kpu
tp:a
tpâ
tpê
gh:ââ
ghuu
lva
lv:ê
lvi
t:êê
); 8 bisyllables (
kamo
kipo
kani
tupa
noki
piwa
towi
nomi);
12 trisyllables (
nademo
meyadi
diyeto
widone
nuyedi
tumowe
pedumi
dimope
tiwune
wumipo
nayeki
mituye); and 4 quadrisyllables (
todiwuma
wadikeno
nomiwake
diponate).

A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets. The 40 elicitation sets are available online from https://osf.io/5qspb. The split had the following constraints:

- the same three items were selected as practice items and used in all 40 elicitation sets
- splits were done within each length group from the 2018 items (i.e., separately for 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists). Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
- once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2_syllables, then 3-syllables, and finally 4-syllable items.

### Procedure

@@

Indeed, we visited four different hamlets once, and attempted to test all eligible childr en present at that time, to prevent the items 'spreading' through hearsay. In the first village, we tested children in five different places, with some children being tested inside their house and others tested on the veranda of a house even if it was not their own. The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.

The child was donned a headset (xx @mc please fill in for most of the children, xx @ac fill in for the rest). For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth. A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything. An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.

@@

### Coding

@@

We sought additional information by asking a local informant to listen through each child production, paired with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated. We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets). This "native coding" provides both accuracy and qualitative desciptions of what the child said that are likely more reliable than those of the non-native coder.

### Analyses

For NWR accuracy, we considered separately the child's first and subsequent attempts. These were scored as correct or incorrect on the fly depending on whether the native experimenter asked for the item to be repeated; as well as offline. Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy [CITE]. We calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt). Although previous work does not use distance metrics, we additionally report those.

Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account. We describe the proportion of items where the change resulted in a real word (semantic errors); and classify the most common phonological errors.

### Participants

@@

read.csv("NWR-demo.csv",header=T)->demo
#colnames(demo)

#if family is missing, attribute each child to a different one
demo$familyID=as.character(demo$familyID)
demo$familyID[is.na(demo$familyID)]<-paste("k",demo$ID[is.na(demo$familyID)])

subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
#if age.exact is missing, use reported age
inc$age.exact[is.na(inc$age.exact)]<-inc$Age[is.na(inc$age.exact)]
#summary(inc)
table(inc$birthOrder)->mytab
```

A total `r dim(demo)[1]` children were tested, from `r length(levels(factor(demo$familyID)))` of families, in five hamlets. Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xd5t repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`). In addition, `r sum(demo$Age >12 | demo$included=="too old",na.rm=T)` teenagers were tested to put younger children at ease; their data is not included in analyses below. The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.exact),2)` years (range `r round(min(inc$age.exact),2)`-`r round(max(inc$age.exact),2)` years). There were `r sum(inc$monolingual=="yes", na.rm=T)` children exposed only to Yélî in the home, `r sum(inc$monolingual=="no", na.rm=T)` children who were also exposed to another language in the home, and , `r sum(is.na(inc$monolingual))` for whom this information was missing. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years; `r sum(is.na(inc$matEd))` children had this information missing). In terms of birth order, `r mytab[1]` were first borns, `r mytab[2]` second,  `r mytab[3]` third,  `r mytab[4]` forth,  `r mytab[5]` fifth, `r mytab[6]` sixth, and  `r mytab[7]` did not have this information.

@mc why is maternal education always in even years? Can you add a footnote to explain?

## Results

- What is the overall repetition accuracy (whole word, phoneme based, distance)?

- How does this change as a function of item complexity (number of syllablex, sound complexity)?

- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?

- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?

```{r}
read.table("log2.txt",header=T,sep="\t")->x
x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
x[order(x$out),]->x
x$cor.online=1
for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0

@@

plot(ms$score.online~ms$score)
```

@@

Based on previous work, one can make the following general predictions for any population:

1. Children are more accurate for mono-syllables than longer items
2. Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds [CITE]. Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
3. Children's accuracy increases with child age.
4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language) 
<!-- **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)** ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)-->
5. As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children

@@

- What is the overall repetition accuracy (whole word, phoneme based, distance)?
- How does this change as a function of item complexity (number of syllablex, sound complexity)?
- How frequent are errors that result in real words? Is that a function of item complexity?
- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
